---
title: "ARA 669 - Assignment 2"
author: "Abdoulie B. Jallow"
date: "today"
format:
  html:
    theme: flatly  # Other theme names include cosmo, lux, or sandstone
    toc: true # Adds an automatic Table of Contents
    toc-depth: 3 # Navigation lists headings down to sub-subheadings
    toc-location: left # Places the navigation menu on the left side
    number-sections: true # Automatically adds numbers (1., 1.1) to your headings
    code-fold: true # Collapses code into neat, clickable boxes
    code-summary: "Show Code" # Sets the text displayed on the clickable code box
    linkcolor: "#18bc9c" # Changes hyperlinks to a clean coastal teal color
    code-block-bg: "#f8f9fa" # Gives code boxes a soft off-white background
    code-block-border-left: "#2c3e50" # Adds a dark slate accent line to code boxes
execute:
  warning: false # Hides messy package warning alerts in your final report
  message: false # Hides general package loading text in your final report
---

## Overview

The goal of this assignment is to apply the tools and approaches we've learned during lab to explore data from a study of periwinkle snail (*Littoraria irrorata*) association with big cordgrass (*Spartina cynosuroides*) and smooth cordgrass (*Spartina alterniflora*) in a salt marsh in Taskinas Creek, a tributary of the York River. The dataset includes the following variables:

obs: Observation number
quad: Quadrat number
plantType: SPAL = *Spartina alterniflora*; SPCY = *Spartina cynosuroides*
liveStems: Number of live stems
deadStems: Number of dead stems
burrows: Fiddler crab burrows greater than one centimeter
mussels: Ribbed mussels (*Guekensia demissa*)
snails: Periwinkle snails (*Littoraria irrorata*)
snailDensity: Snail Density
notes: any notes about the observation

*Please submit this assignment via GitHub by 24 September.*  


## Getting started

**Use the code chunk below to load packages and the data. \[0.5 pt\]**

```{r setup}
#| message: false
#| error: false

# Paste your code in here!
# required packages
library(tidyverse)
library(readr) 

# Data importation and data structure 

data <- read_csv("Failon_Plant_Stem_Densities_and_Invertebrate_Survey.csv")
summary(data)
dim(data)
str(data)
attach(data)


## theme setting

theme_set(
  theme_classic(base_size = 12) + 
    theme(
      plot.title    = element_text(size = 15, face = "bold", hjust = 0.5, margin = margin(b = 10)),
      plot.subtitle = element_text(size = 11, face = "italic", hjust = 0.5, margin = margin(b = 10)),
      axis.title    = element_text(size = 12, face = "bold"),
      axis.text     = element_text(size = 10, color = "black"),
      legend.title  = element_text(size = 11, face = "bold")))

```

## Data exploration

**Explore the data visually using ggplot2. \[0.5 pt\]**
I Choose the liveStems, burrows and deadstems as my predictors for snail density, before I went ahead with my models I plotted histograms of these 3 variables to see how distributed are they before going ahead with my models. I use used ggpairs on the 3 variables to see how well the correlated the outcome/response(SnailDensity) but most of them were poorly correlated with the response variable. Burrows and deadstemps shows a negative co-relationship while livestems show very minimal corelatinship (0.035)

```{r plotting}
#| message: false
#| error: false

library(GGally)
library(patchwork) # place all the plots side side 
library(DataExplorer)

dat <- data %>% 
  select(3:8) %>% 
  plot_intro() # visualize data types and missing percentages variables


#lattice::splom(data)


#looking at the histogram of the variables
#par(mfrow = c(1,3)) # plot all the three

#hist(liveStems, main = "liveStems", xlab = "liveStems", col = "lightblue")

#hist(burrows, main = "burrows", xlab = "burrows", col = "lightgreen")

#hist(deadStems, main = "Deadstems", xlab = "Deadstems", col = "lightpink")


##GGPlot ploting diagnostic plots
library(ggplot2)
library(patchwork)

Q1 <- ggplot(data.frame(liveStems), aes(x = liveStems)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 20) +
  theme_classic() +
  labs(title = "Live Stems")
Q2 <- ggplot(data.frame(burrows), aes(x = burrows)) +
  geom_histogram(fill = "lightgreen", color = "black", bins = 20) +
  theme_classic() +
  labs(title = "Burrows")

Q3<- ggplot(data.frame(deadStems), aes(x = deadStems)) +
  geom_histogram(fill = "lightpink", color = "black", bins = 20) +
  theme_classic() +
  labs(title = "Dead Stems")




#GGally::ggpairs(data)

data |> 
select(snailDensity, liveStems, deadStems, burrows) %>%
ggpairs()


#ggplot(data, aes(mussels, snailDensity)) + geom_point() + geom_smooth(method = "lm")
    
#ggplot(data, aes(x = plantType, y = snailDensity)) + geom_boxplot()


#Q1+ Q2+ Q3 #Using the patchwork package




#data |> select(,3:8) |> ggpairs()

plot1 <- ggplot(data = data, aes(x= liveStems,y=snailDensity)) + 
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm",se = TRUE) +
    labs(
      title = "Snail Density vs LiveStems",
      subtitle = "Testing the relationship between liverstems and sanil density",
      x="LiveStems",
      y= "Snail Density")



plot2 <- ggplot(data = data, aes(x = deadStems, y = snailDensity)) + 
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Snail Density vs Dead Stems",
    subtitle = "Testing the relationship between Deadstems and Snaildensity",
    x = "Dead Stems",
    y = "Snail Density ")



plot3 <- ggplot(data = data, aes(x = burrows, y = snailDensity)) + 
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Snail Density vs Burrows",
    subtitle = "Testing the relationship between burrows and snail density",
    x = "Burrows ",
    y = "Snail Density")



# Place all three plots side-by-side in 1 row and 3 columns
library(patchwork) # place all the plots side side 
plot1 + plot2 + plot3

```




**Describe your approach. What choices did you make in deciding which variables to use, plot types, etc., and why? \[1 pt\]**
I explored my chosen variables which are the livestems, deadStems and burrows,  using histograms. Base on the histogram of plot 1, plot2 and plot3 I would say they are normally distributed. I choosed these variables as my predictors due to the fact that I think there will have an effect on the snail density and might help to answer the question about snail variability. 

## Research question

**Based on the data description above and your visualizations, formulate research questions and testable hypotheses that could be answered by applying linear models to the data provided. \[1 pt\]**

I will evaluate three separate simple linear regression models to determine which individual environmental factor best predicts snail density.

### Model 1: Live Stems Hypothesis

**The number of live stems has no effect on snail density **.

### Model 2: Dead Stems Hypothesis

**The number of dead stems has no effect on snail density**.

### Model 3: Burrows Hypothesis
**The number of burrows has no effect on snail density**.

## Analysis

**Develop a set of models (3+) that reflect the hypotheses described above. \[1 pt\]**

```{r models}
#| message: false
#| error: false

# Paste your code in here!
m1 <-lm(snailDensity~liveStems,data=data)
m2 <-lm(snailDensity~liveStems+ burrows,data=data)
m3 <- lm(snailDensity~liveStems+ burrows+deadStems ,data=data)
m4<- lm(snailDensity~liveStems+ burrows*deadStems ,data=data) # assessing interaction term
m5<- lm(snailDensity~liveStems+ burrows*liveStems,data=data)

summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)

```

**Evaluate your models using an approach covered as part of this class. \[1 pt\]**
I explored 5 model to see which best explained the snaildensity. Models of m1 to M3 was to assess significant relationship and models of m4 and m5 was to test intraction between burrows and deadstems and burrows and livestems. Base on the DeltaAIC scores model(m2) is the best. However in all of them only burrows shows significant relationship to the SnailDensity

```{r modelselection}
#| message: false
#| error: false

# Paste your code in here!

AIC(m1, m2, m3, m4,m5) |> 
  arrange(AIC) |> 
  mutate(deltaAIC = AIC - min(AIC))

```

**Check your model diagnostics (i.e., are the model assumptions met) using an approach covered as part of this class. \[1 pt\]**
looking at the residual plots, variables livestems and deadStems violates the assumption rule while burrows satisfy the assumptions by showing linearity.  

```{r diagnostics}
#| message: false
#| error: false

# Paste your code in here!

library(performance)

car::crPlots(m2)     # from the car package 

performance::check_model(m2)
sjPlot::plot_model(m2, type = "diag")

#Model assumptions checl
#plot(m3, which = 1) # residual vs plotted 
#plot(m3, which = 2)  # qqplot for nomality check
qqnorm(residuals(m2))
#qqline(residuals(m3)) # did not execute 

#plot(m3, which = 3)   # scale location plot
#plot(m3, which = 4)   # cook's distance 

performance::check_outliers(m2) ## outliers check 
performance::check_collinearity(m2)  # multicolinearlity check
car::vif(m2)


```

**Provide parameter estimates and SEs for the most supported model and plot the fitted model over observed data. \[2 pts\]**
My m2 model was the most supported one which I have chosen one at the end of exploring 5 models. Model m2 estimated an intercept of 180.355 with a SE of 25.876 while livestems have an estimated effect of 1.245 with SE of 1.813, Burrows had estimated effect of -14.596 with SE = 5.127


```{r}

# Plot 1: Burrows
p1 <- ggplot(data, aes(x = burrows, y = snailDensity)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Snail Density by Burrows")

# Plot 2: Live Stems
p2 <- ggplot(data, aes(x = liveStems, y = snailDensity)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(title = "Snail Density by Live Stems")

# Display them side-by-side using patchwork syntax
p1 + p2


#data %>% select(snailDensity, burrows, liveStems) %>% pivot_longer(cols = c(burrows, liveStems), 
               #names_to = "Predictor", 
               #values_to = "Value") %>%

  #ggplot(aes(x = Value, y = snailDensity)) + geom_point(size = 2) + geom_smooth(method = "lm", se = TRUE, color = "blue") +
  # This splits the plot into two panels based on the predictor type
  #facet_wrap(~ Predictor, scales = "free_x") 

# confident intervals of m3
confint(m2)

```


## Conclusions

**Provide summary statements about the quality of your analysis, including any potential limitations or caveats. \[1 pt\]**
I ran 5 models hoping to see meaningful biological or ecological explanation of Snail density but only burrows was statistically significant  linear relationship with snail density. I was hoping to see the interactive effect between deadstem and burrows there was non too. The limitation is that the model assumes linear relationships which was not the case as shown by the component-plus-residual plots of livestems and deadstems. 

## AI statement

*If you used AI for this assignment, please describe your use, including:*
I use copilot to interpret the interact effect and my check model performance did display the plots show explore under options using copilot. 

*1.  What tool or service did you use?*
I use copilot 

*2.  What was the specific purpose or use?*
to interpret the interactive effect 

*3.  How did this use enhance your understanding of the topic/code?*
i use it also to evaluate the outcomes of my models wether i did a bad choice picking these varibales. 

## Final Project

Each student will complete an independent project using a dataset from their own research whenever possible. Students who do not have an appropriate dataset will be provided with one. **If you do not have a dataset to use for this project, please reach out to Rob & Gina as soon as possible.**

The project is intended to mirror the statistical component of a scientific research study and will require students to integrate concepts learned throughout the course. Project components include: 

- formulation of a research question, objective, and/or testable hypotheses;
- description of the study design and data with justification of design adequacy;
- exploratory data analysis;
- selection and justification of an appropriate modeling approach;
- model fitting, comparison, selection, and validation;
- interpretation of results in the context of the research question;
- preparation of publication-quality figures and tables;
- a written report in the style of a short scientific paper (approximately 4-6 pages); and
- a 3-minute lightening talk summarizing the study, analytical approach, results, and conclusions.

Projects will be evaluated on the quality of the scientific question, appropriateness of the statistical methodology, rigor of the analysis, interpretation of results, clarity of written and oral communication, and reproducibility of the analytical workflow.

To support you on this analytical journey, this project has been divided into four steps, which will all contribute to your final project grade (worth 50% of your total grade). 

1. Project proposal, including research question and description of dataset
2. Exploratory data analysis and proposed analytical approach
3. Preliminary model fits and diagnostics
4. Final paper and presentation

## Project proposal


**Formulate 1-2 research questions, objectives, and/or testable hypotheses of interest. Provide sufficient detail for someone not familiar with your work to understand the goal(s) of your project. \[2 pts\]**



**Describe the study design and data, including justification of design adequacy to address the research question, objectives, and/or testable hypotheses described above. Provide sufficient detail to evaluate your data in the context of your questions of interest. \[3 pts\]**



